\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-10-18 doc/paper-jss/sdmTMB-paper.Rnw
Wrap paragraph
Commands

\Abstract{
Geostatistical data are common across scientific fields and the dimensions of these data continue to become larger in space and time.
However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging for many users to implement.
Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation).
Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation.
Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \pkg{sdmTMB} can be an order of magnitude faster than \proglang{R}-\pkg{INLA}.
We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
}

@@

Such models quickly become computationally challenging due to the need to invert large matrices to account for covariation among data.
Many solutions have been proposed, such as predictive processes \citep{banerjee2008, latimer2009}, the stochastic partial differential equation (SPDE) approximation to GRFs \citep{lindgren2011}, and nearest-neighbour Gaussian processes \citep{datta2016, finley2022}.
These approaches aim to minimize the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian \citep{gelfand2017} or maximum likelihood methods.
This can greatly improve computational efficiency \citep[e.g.,][]{heaton2019}.
The SPDE approach has been widely adopted, especially via the \proglang{R}-\pkg{INLA} package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \pkg{TMB} \citep[Template Model Builder,][]{kristensen2016} that relies on \proglang{R}-\pkg{INLA} to create input matrices \citep[e.g.,][]{thorson2019, osgood-zimmerman2021}.

@@

Software packages designed for specifying statistical models which incorporate the SPDE, such as \proglang{R}-\pkg{INLA} and \pkg{TMB}, are flexible and powerful but can be challenging for many applied researchers.
For example, \pkg{TMB} requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code.
While packages such as \pkg{lme4} \citep{bates2015} and \pkg{glmmTMB} \citep{brooks2017} let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---they lack built-in SPDE or spatiotemporal functionality.
Packages such as \pkg{VAST} \citep{thorson2019} and \pkg{inlabru} \citep{bachl2019} are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, or lack some functionality.

@@

Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \INLA\ to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
Many features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: \pkg{sdmTMB}), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and discretely in time.

This paper describes the underlying statistical models (Section~\ref{model-description}), explains how \pkg{sdmTMB} is designed and summarizes its software functionality (Section~\ref{sec:software}), illustrates its use through three case studies (Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}), compares \pkg{sdmTMB} to other packages (Section~\ref{package-comparisons}) and concludes with a discussion of links to other packages and future development directions (Section~\ref{discussion}).

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples, notation, and example code.
An \pkg{sdmTMB} model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation: We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See the \pkg{sdmTMB} \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint `hockey-stick' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}

@@

% \pkg{sdmTMB} fits GLMMs to spatial or spatiotemporal geostatistical data.
% Geostatistical data refers to data observed at specific spatial coordinates
% reflecting some underlying spatial process \citep{rossi1992, diggle2007}. These
% data can be collected across discrete points in time. The process component of
% an \pkg{sdmTMB} model can be formed by any combination of main (``fixed'')
% effects \(\bm{\beta}\), an offset \(O_{\bm{s},t}\), IID (independent and
% identically distributed) random intercepts \(\alpha_g\), spatial intercept
% random fields \(\omega_{\bm{s}}\), spatiotemporal intercept random fields
% \(\epsilon_{\bm{s},t}\), time-varying effects \(\bm{\gamma_t}\), and spatially
% varying effects \(\zeta_{\bm{s}}\) (Figure~\ref{fig:diagram}). These components
% are discussed in more detail via our examples below.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=3.4in]{flowchart}
\caption[]{
  Description of the model fitting procedure in \pkg{sdmTMB}.
}\label{fig:flowchart}
\end{figure}

@@

Once the user makes a call to \texttt{sdmTMB()}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\texttt{sdmTMB()} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.

@@

\pkg{sdmTMB} models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \texttt{sdmTMB()} \texttt{prior} argument (\texttt{?sdmTMBpriors}).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \pkg{TMB} \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\texttt{?pc\_matern}) can constrain the spatial range and variance parameters.
These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
If a model will be passed to \pkg{tmbstan} with priors, a \texttt{bayesian} logical argument should be set to \texttt{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.

@@

% The \pkg{sdmTMB} model is fit by maximum marginal likelihood. Internally,
% a \pkg{TMB} \citep{kristensen2016} model template is used to calculate the
% marginal log likelihood and its gradient, and the negative log likelihood is
% minimized via the non-linear optimization routine \texttt{stats::nlminb()} in
% \proglang{R} \citep{gay1990, r2021}. Random effects are estimated at values that
% maximize the log likelihood conditional on the estimated fixed effects and are
% integrated over via the Laplace approximation \citep{kristensen2016}. After
% rapid model exploration with maximum likelihood, one can optionally pass an
% \pkg{sdmTMB} model to the \proglang{R} package \pkg{rstan} \citep{carpenter2017,
% rstan} via \pkg{tmbstan} \citep{monnahan2018} to estimate the joint posterior
% distribution for Bayesian inference.

@@

% Validation and selection of state-space models is challenging, particularly when
% using the Laplace approximation \citep{thygesen2017a}. We provide several
% approaches to assist this process: (1) The Akaike Information Criterion
% \citep[AIC,][]{akaike1974} can be calculated with \texttt{AIC()}, although AIC
% has well-documented biases with mixed-effects models \citep{liang2008}. (2)
% Alternatively, k-fold cross validation with \texttt{sdmTMB\_cv()} can be used
% with user-specified or randomly chosen folds for model selection or to evaluate
% goodness of fit according to user-calculated criteria (e.g., mean squared error,
% area under the curve). (3) An \pkg{sdmTMB} model can be passed to the
% \pkg{tmbstan} package \citep{monnahan2018} to sample from the joint posterior
% with Stan \citep{carpenter2017}, evaluate the accuracy of the Laplace
% approximation, or perform posterior predictive checks (see
% \texttt{?extract\_mcmc}). (4) The \texttt{residuals()} method by default returns
% randomized quantile \citep{dunn1996} or probability integral transform (PIT)
% \citep{smith1985} residuals. For state-space models, these residuals have known
% statistical issues with the Laplace approximation \citep{thygesen2017a} but are
% quick to calculate. A version that uses Markov chain Monte Carlo (MCMC) sampling
% of the random effects to avoid this issue \citep{rufener2021} is recommended but
% slower (\texttt{?residuals.sdmTMB}). Simulation-based residuals \citep{dharma}
% are also possible. (5) The \texttt{simulate.sdmTMB()} method can simulate from
% fitted models and the \texttt{sdmTMB\_simulate()} function facilitates
% simulating data without starting from a fitted model. Models can be fit to these
% simulated data to ensure identifiability, evaluate bias and precision in
% parameter estimation, or evaluate the consequences of model misspecification.

@@

% \sdmTMB\ relies on several well-established \R\ packages (Figure~\ref{fig:flowchart}).
% \sdmTMB\ uses \pkg{fmesher} \citep{fmesher} and \INLA\ \citep{lindgren2015} to generate the triangulation mesh and input matrices for the SPDE calculations.
% The SPDE precision matrix calculations themselves are performed by functions within \pkg{TMB} \citep{kristensen2016} (\texttt{R\_inla} namespace).
% Data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
% The statistical model itself is written as a \proglang{C++} template for \TMB.
% \TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
% The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
% After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
% % TODO: visreg + ggeffects effects, etc.?

@@

\pkg{sdmTMB} can be installed from the Comprehensive \proglang{R}
Archive Network (CRAN) at
\url{https://CRAN.R-project.org/package=sdmTMB}

@@

We begin with a simple species distribution model of encounter probability of Pacific Cod (\emph{Gadus macrocephalus}) from a trawl survey conducted in Queen Charlotte Sound, British Columbia, Canada.
The purpose of our example is to illustrate the need for spatial random fields.
This survey is conducted by Fisheries and Oceans Canada and follows a depth-stratified random sampling design, resulting in a georeferenced dataset.
The data frame \texttt{pcod} is available as package data in \pkg{sdmTMB}.
Relevant columns include latitude, longitude, Universal Transverse Mercator (UTM) coordinates, bottom depth, and encounter (\texttt{present = 1}) vs.~non-encounter (\texttt{present = 0}) of Pacific Cod for a given survey sample.

@@

An \pkg{sdmTMB} model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
Usually it makes sense to convert the spatial coordinates to an equidistant projection such as UTMs to ensure that distance remains constant throughout the study region \citep[e.g., using \texttt{sf::st\_transform()},][]{pebesma2018}.
Here we use the helper function \texttt{add\_utm\_columns()} to add UTM coordinates with km units (so our estimated spatial range parameter is not too big or small).
By default, the function will guess the UTM zone and create new columns \texttt{X} and \texttt{Y}.
Since our example data already has these UTM columns, we can skip running this code.

@@

We can also specify additional arguments, in this case passed to \texttt{fmesher::fm\_mesh\_2d\_inla()}: a maximum triangle edge length of 10 km and 40 km for the inner and outer mesh, and an offset width of 10 km and 40 km for the inner and outer meshes borders.
% Alternatively, we could have created any mesh via \proglang{R}-\pkg{INLA} or \pkg{fmesher} and supplied it to the \texttt{mesh} argument in \texttt{make\_mesh()}.
For more irregularly shaped areas, we could have specified a non-convex hull with \texttt{convex} and \texttt{concave} arguments.
The triangle edge length should be several times smaller than the range and the outer boundary should extend at least as far as the range (REF).

@@

As a second example, we will construct a spatiotemporal model of catch rates of Pacific Spiny Dogfish (\emph{Squalus suckleyi}) from a trawl survey off the west coast of Vancouver Island, Canada.
This example extends the spatial model described above by also including (1) spatiotemporal fields allowing the distribution in each year to be unique, (2) a time-varying intercept as an AR(1) process to allow year effects to vary but remain autocorrelated, (3) a smooth effect of depth, allowing catch rates to vary non-linearly with depth, and (4) spatial anisotropy allowing spatial correlation to be directionally dependent.
Since catch rates are positive, continuous, and contain zeros, we begin by specifying the response family as a Tweedie distribution \citep{tweedie1984} with a log link.
We then compare alternative families, spatiotemporal random field structures, and the exclusion of anisotropy to illustrate the flexibility of \pkg{sdmTMB}.

@@

There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models \citep[e.g.,][]{heaton2019}.
\pkg{sdmTMB}, \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} \citep{rousset2014} are the most closely related, as they all provide a user interface to SPDE-based GMRF models.
In our software comparison (Table \ref{tab:functionality}), we also include \pkg{mgcv} as it can be adapted to use the SPDE \citep{miller2019} and \pkg{spBayes} \citep{finley2007, finley2015} since it is a prominent package that can fit related predictive-process models without the SPDE.
\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
To our knowledge, \pkg{VAST} is the only package to implement spatial \citep{thorson2015} and spatial dynamic factor analysis \citep{thorson2016} and spatial empirical orthogonal function (EOF) regression \citep{thorson2020}.
Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
\pkg{spaMM} is limited to a spatial random field (i.e., does not fit spatiotemporal fields) and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
% We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}, and \pkg{inlabru} in Appendix~\ref{app:inla}.

@@

\subsection[Example sdmTMB model fit]{Example \pkg{sdmTMB} model fit}

>>>>

\Abstract{
Geostatistical data are common across scientific fields and the dimensions of these data continue to become larger in space and time.
However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging for many users to implement.
Here, we introduce the \R\ package \sdmTMB, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
SPDE matrices are constructed with \INLA\ and estimation is conducted via maximum marginal likelihood with Template Model Builder (\TMB) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
We describe the model and explore case studies that illustrate \sdmTMB's flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation).
Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation.
Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \sdmTMB\ can be an order of magnitude faster than \INLA.
We hope \sdmTMB's accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
}

@@

Such models quickly become computationally challenging due to the need to invert large matrices to account for covariation among data.
Many solutions have been proposed, such as predictive processes \citep{banerjee2008, latimer2009}, the stochastic partial differential equation (SPDE) approximation to GRFs \citep{lindgren2011}, and nearest-neighbour Gaussian processes \citep{datta2016, finley2022}.
These approaches aim to minimize the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian \citep{gelfand2017} or maximum likelihood methods.
This can greatly improve computational efficiency \citep[e.g.,][]{heaton2019}.
The SPDE approach has been widely adopted, especially via the \INLA\ package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \TMB\ \citep[Template Model Builder,][]{kristensen2016} that relies on \INLA\ to create input matrices \citep[e.g.,][]{thorson2019, osgood-zimmerman2021}.

@@

Software packages designed for specifying statistical models that incorporate the SPDE, such as \INLA\ and \TMB, are flexible and powerful but can be challenging for many applied researchers.
For example, \TMB\ requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code.
While packages such as \pkg{lme4} \citep{bates2015} and \pkg{glmmTMB} \citep{brooks2017} let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---they lack built-in SPDE or spatiotemporal functionality.
Packages such as \pkg{VAST} \citep{thorson2019} and \pkg{inlabru} \citep{bachl2019} are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, or lack some functionality.

@@

Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \TMB\ for model fitting and \INLA\ to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
Many features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB: \sdmTMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and discretely in time.

This paper describes the underlying statistical models (Section~\ref{model-description}), explains how \sdmTMB\ is designed and summarizes its software functionality (Section~\ref{sec:software}), illustrates its use through three case studies (Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}), compares \sdmTMB\ to other packages (Section~\ref{package-comparisons}) and concludes with a discussion of links to other packages and future development directions (Section~\ref{discussion}).

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{../../figs/diagram}
\caption[]{
Components of an \sdmTMB\ model with illustrations, descriptions, examples, notation, and example code.
An \sdmTMB\ model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation: We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See the \sdmTMB\ \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint `hockey-stick' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in an \R\ formula to omit an intercept.}\label{fig:diagram}
\end{figure}

@@

% \sdmTMB\ fits GLMMs to spatial or spatiotemporal geostatistical data.
% Geostatistical data refers to data observed at specific spatial coordinates
% reflecting some underlying spatial process \citep{rossi1992, diggle2007}. These
% data can be collected across discrete points in time. The process component of
% an \sdmTMB\ model can be formed by any combination of main (``fixed'')
% effects \(\bm{\beta}\), an offset \(O_{\bm{s},t}\), IID (independent and
% identically distributed) random intercepts \(\alpha_g\), spatial intercept
% random fields \(\omega_{\bm{s}}\), spatiotemporal intercept random fields
% \(\epsilon_{\bm{s},t}\), time-varying effects \(\bm{\gamma_t}\), and spatially
% varying effects \(\zeta_{\bm{s}}\) (Figure~\ref{fig:diagram}). These components
% are discussed in more detail via our examples below.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=3.4in]{flowchart}
\caption[]{
  Description of the model fitting procedure in \sdmTMB.
}\label{fig:flowchart}
\end{figure}

@@

Once the user makes a call to \texttt{sdmTMB()}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\texttt{sdmTMB()} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.

@@

\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \texttt{sdmTMB()} \texttt{prior} argument (\texttt{?sdmTMBpriors}).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\texttt{?pc\_matern}) can constrain the spatial range and variance parameters.
These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
If a model will be passed to \pkg{tmbstan} with priors, a \texttt{bayesian} logical argument should be set to \texttt{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.

@@

% The \sdmTMB\ model is fit by maximum marginal likelihood. Internally,
% a \TMB\ \citep{kristensen2016} model template is used to calculate the
% marginal log likelihood and its gradient, and the negative log likelihood is
% minimized via the non-linear optimization routine \texttt{stats::nlminb()} in
% \R\ \citep{gay1990, r2021}. Random effects are estimated at values that
% maximize the log likelihood conditional on the estimated fixed effects and are
% integrated over via the Laplace approximation \citep{kristensen2016}. After
% rapid model exploration with maximum likelihood, one can optionally pass an
% \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017,
% rstan} via \pkg{tmbstan} \citep{monnahan2018} to estimate the joint posterior
% distribution for Bayesian inference.

@@

% Validation and selection of state-space models is challenging, particularly when
% using the Laplace approximation \citep{thygesen2017a}. We provide several
% approaches to assist this process: (1) The Akaike Information Criterion
% \citep[AIC,][]{akaike1974} can be calculated with \texttt{AIC()}, although AIC
% has well-documented biases with mixed-effects models \citep{liang2008}. (2)
% Alternatively, k-fold cross validation with \texttt{sdmTMB\_cv()} can be used
% with user-specified or randomly chosen folds for model selection or to evaluate
% goodness of fit according to user-calculated criteria (e.g., mean squared error,
% area under the curve). (3) An \sdmTMB\ model can be passed to the
% \pkg{tmbstan} package \citep{monnahan2018} to sample from the joint posterior
% with Stan \citep{carpenter2017}, evaluate the accuracy of the Laplace
% approximation, or perform posterior predictive checks (see
% \texttt{?extract\_mcmc}). (4) The \texttt{residuals()} method by default returns
% randomized quantile \citep{dunn1996} or probability integral transform (PIT)
% \citep{smith1985} residuals. For state-space models, these residuals have known
% statistical issues with the Laplace approximation \citep{thygesen2017a} but are
% quick to calculate. A version that uses Markov chain Monte Carlo (MCMC) sampling
% of the random effects to avoid this issue \citep{rufener2021} is recommended but
% slower (\texttt{?residuals.sdmTMB}). Simulation-based residuals \citep{dharma}
% are also possible. (5) The \texttt{simulate.sdmTMB()} method can simulate from
% fitted models and the \texttt{sdmTMB\_simulate()} function facilitates
% simulating data without starting from a fitted model. Models can be fit to these
% simulated data to ensure identifiability, evaluate bias and precision in
% parameter estimation, or evaluate the consequences of model misspecification.

@@

% \sdmTMB\ relies on several well-established \R\ packages (Figure~\ref{fig:flowchart}).
% \sdmTMB\ uses \pkg{fmesher} \citep{fmesher} and \INLA\ \citep{lindgren2015} to generate the triangulation mesh and input matrices for the SPDE calculations.
% The SPDE precision matrix calculations themselves are performed by functions within \TMB\ \citep{kristensen2016} (\texttt{R\_inla} namespace).
% Data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
% The statistical model itself is written as a \proglang{C++} template for \TMB.
% \TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
% The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
% After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
% % TODO: visreg + ggeffects effects, etc.?

@@

\sdmTMB\ can be installed from the Comprehensive \R
Archive Network (CRAN) at
\url{https://CRAN.R-project.org/package=sdmTMB}

@@

We begin with a simple species distribution model of encounter probability of Pacific Cod (\emph{Gadus macrocephalus}) from a trawl survey conducted in Queen Charlotte Sound, British Columbia, Canada.
The purpose of our example is to illustrate the need for spatial random fields.
This survey is conducted by Fisheries and Oceans Canada and follows a depth-stratified random sampling design, resulting in a georeferenced dataset.
The data frame \texttt{pcod} is available as package data in \sdmTMB.
Relevant columns include latitude, longitude, Universal Transverse Mercator (UTM) coordinates, bottom depth, and encounter (\texttt{present = 1}) vs.~non-encounter (\texttt{present = 0}) of Pacific Cod for a given survey sample.

@@

An \sdmTMB\ model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
Usually it makes sense to convert the spatial coordinates to an equidistant projection such as UTMs to ensure that distance remains constant throughout the study region \citep[e.g., using \texttt{sf::st\_transform()},][]{pebesma2018}.
Here we use the helper function \texttt{add\_utm\_columns()} to add UTM coordinates with km units (so our estimated spatial range parameter is not too big or small).
By default, the function will guess the UTM zone and create new columns \texttt{X} and \texttt{Y}.
Since our example data already has these UTM columns, we can skip running this code.

@@

We can also specify additional arguments, in this case passed to \texttt{fmesher::fm\_mesh\_2d\_inla()}: a maximum triangle edge length of 10 km and 40 km for the inner and outer mesh, and an offset width of 10 km and 40 km for the inner and outer meshes borders.
% Alternatively, we could have created any mesh via \INLA\ or \pkg{fmesher} and supplied it to the \texttt{mesh} argument in \texttt{make\_mesh()}.
For more irregularly shaped areas, we could have specified a non-convex hull with \texttt{convex} and \texttt{concave} arguments.
The triangle edge length should be several times smaller than the range and the outer boundary should extend at least as far as the range (REF).

@@

As a second example, we will construct a spatiotemporal model of catch rates of Pacific Spiny Dogfish (\emph{Squalus suckleyi}) from a trawl survey off the west coast of Vancouver Island, Canada.
This example extends the spatial model described above by also including (1) spatiotemporal fields allowing the distribution in each year to be unique, (2) a time-varying intercept as an AR(1) process to allow year effects to vary but remain autocorrelated, (3) a smooth effect of depth, allowing catch rates to vary non-linearly with depth, and (4) spatial anisotropy allowing spatial correlation to be directionally dependent.
Since catch rates are positive, continuous, and contain zeros, we begin by specifying the response family as a Tweedie distribution \citep{tweedie1984} with a log link.
We then compare alternative families, spatiotemporal random field structures, and the exclusion of anisotropy to illustrate the flexibility of \sdmTMB.

@@

There are many \R\ packages capable of fitting geostatistical spatial or spatiotemporal models \citep[e.g.,][]{heaton2019}.
\sdmTMB, \pkg{VAST}, \INLA/\pkg{inlabru}, and \pkg{spaMM} \citep{rousset2014} are the most closely related, as they all provide a user interface to SPDE-based GMRF models.
In our software comparison (Table \ref{tab:functionality}), we also include \pkg{mgcv} as it can be adapted to use the SPDE \citep{miller2019} and \pkg{spBayes} \citep{finley2007, finley2015} since it is a prominent package that can fit related predictive-process models without the SPDE.
\sdmTMB, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \INLA/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
\sdmTMB\ and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \INLA/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
To our knowledge, \pkg{VAST} is the only package to implement spatial \citep{thorson2015} and spatial dynamic factor analysis \citep{thorson2016} and spatial empirical orthogonal function (EOF) regression \citep{thorson2020}.
Of these packages, only \sdmTMB\ and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
\pkg{spaMM} is limited to a spatial random field (i.e., does not fit spatiotemporal fields) and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
% We provide comparisons of the syntax and the reproducibility of results from models fit using \sdmTMB, \INLA, and \pkg{inlabru} in Appendix~\ref{app:inla}.

@@

\subsection[Example sdmTMB model fit]{Example \sdmTMB\ model fit}
